### 2016-08-01 part1/lafaySimulatedScenePerception1.tex
udpate NM comments

\newcommand{\gl}[1]{\textcolor{blue}{GL : #1}}
\newcommand{\todo}[1]{\textcolor{red}{todo : #1}}
\newcommand{\ml}[1]{\textcolor{magenta}{ML : #1}}
\newcommand{\mr}[1]{\textcolor{green}{MR : #1}}

@@

As a case study to demonstrate the interest of that new approach, we focus in this paper on the issue of urban soundscape pleasantness as tackled in \cite{guastavino_ideal_2006} \gl{NM comments}.

@@

We first give an overview of the tools used in soundscape perception studies, and justify the benefits of conducting this kind of simulation experiment as a data generation process. We then describe the simulation process, and present the results of the case study. The latter consists of two experiments: 1) a simulation experiment where subjects generate both ideal and non-ideal urban sonic environments, and 2) a pleasantness evaluation of the simulated scenes using a 7 points bipolar semantic scale \gl{NM comments}.

Results show that \gl{NM comments} 1) outcomes identified in the literature \cite{guastavino_categorization_2007} can be achieved by this approach, providing evidence that it has solid grounding, 2) sound markers that greatly affect the pleasantness of the soundscape can be identified with a good level of categorical detail, and 3) that physical descriptors computed specifically on the sound markers better characterise the appraisal than global descriptors.

@@

Two approaches, named in this paper \textit{categorical} and \textit{dimensional}, are commonly used to gain knowledge on soundscape perception \gl{NM comments}.

The \textit{categorical} approach investigates both the existence of 1) soundscape categories and 2) soundscape components (categories of sound sources) using both sorting \cite{maffiolo_caracterisation_1999, guastavino_categorization_2007} and descriptive \cite{axelsson2005soundscape, raimbault_urban_2005, raimbault2006qualitative, guastavino_ideal_2006} tasks (questionnaires or interviews). To identify categories, those methods usually require psycho-linguistic and lexical analyses. Their results tend to show that soundscape perception relies upon the identification and assessment of the sound sources which compose the soundscape \cite{dubois2006cognitive}. Maffiolo \cite{maffiolo_caracterisation_1999} shows the existence of two generic soundscape categories, namely the ``\,event sequences\,'' and the ``\,amorphous sequences\,''. \gl{This distinction is also observed in \cite{guastavino_ideal_2006}}. Guastavino \cite{guastavino_categorization_2007} as well as Niessen \emph{et al.} \cite{niessen_categories_2010}  show the prominence of specific sound source categories in soundscape perception: ``\,natural\,'', ``\,human\,'', ``\,traffic\,'' and ``\,mechanical\,''. Other studies not using \gl{NM comments} the \textit{categorical} approach reveal that sound source identification underlines the perception of others factors, for example Kuwano \emph{et al.} \cite{kuwano_memory_2003-1} show that overall judgment of loudness of a sound environment sequence is not statistically different from average instantaneous judgment of the recalled sound events of the sequence \gl{NM comments}.  As reported by \cite{davies_perception_2013}, while the categorical approach does help to understand how a soundscape is composed, it does not provide information on how soundscape components are perceived and influence the qualitative evaluation.

Instead of asking subjects to break soundscapes down into categories, the \textit{dimensional} approach makes use of bipolar semantic scales to describe the soundscape either in a holistic way \cite{jeon2013soundwalk, kang_semantic_2010, cain_development_2013, torija2013application} or by considering the influence of specific sound sources \cite{guyot2005urban, lavandier2006contribution, ricciardi2015sound}. From those descriptors, it is possible to use dimensional analysis techniques such as Principal Component Analysis to find relevant perceptual axes \cite{cain_development_2013}, or to connect those perceptive outcomes with objective acoustics measurements using various statistical analysis or modelling techniques such as multiple linear regression \cite{lavandier2006contribution,ricciardi2015sound} or artificial neural network \cite{yu2009modeling}. Several semantic scales show promising results when it comes to modeling perceived affective quality, namely ``\,Calmness\,'' and ``\,Vibrancy\,'' \cite{cain_development_2013, axelsson2010principal}, or ``\,Pleasantness\,'' and ``\,Eventfulness\,'' \cite{axelsson2010principal}; but it has been shown that others context dependent (``\,Appropriateness\,'',``\,Familiarity\,'') or music related (``\,Music-likeness\,'' ) scales provide useful complementary information on soundscape perception \cite{aletta2016soundscape}. In addition, several studies have shown that multiple factors such as physical, behavioral, social, demographical, and psychological are involved when evaluating acoustic comfort \cite{marquis2005noise, yu2009modeling}. \gl{In particular} \cite{yu2010factors} shows that age and education level significantly influence the sound preference \gl{NM comments, attention \cite{marquis2005noise} est une review}.

\gl{Recent} studies investigating the influence of sound sources in soundscape perception show that global sound level (perceived or measured) cannot entirely explain the quality of a sonic environment \gl{NM comments, plus développer, ajout article ?}. \cite{guyot2005urban} shows that a rise in soundscape quality between two street soundscapes having a similar time-ratio of traffic sounds and a similar global sound level may result from the addition of a positively connotated sound source, e.g. belonging to the ``\,human\,'' category \gl{NM comments: eg means exempli gratia (par exemple)}. Using multiple linear regression to predict unpleasantness from descriptors characterizing global sound level as well as time presence and sound level of each source, \cite{lavandier2006contribution} shows that the percentage of explained variance increases when the source descriptors are added to the global sound level. Similar findings are made in \cite{ricciardi2015sound}, showing that a combined use of global and source specific descriptors allows to build more relevant model of sound quality. Lastly, \cite{lavandier2006contribution} shows that global sound level is not always appropriate to characterize the appraisal, especially in the case of positively rated soundscapes such as \textit{park} or \textit{market}.

In order to be able to study the relative perceptual importance of each sound source, it is beneficial to first gain knowledge about the presence and level of those sound sources within \gl{inside ?} the scene. Acquiring such data may be a serious challenge. From most of the studies cited above \cite{guyot2005urban, lavandier2006contribution, ricciardi2015sound}, subjects are asked to evaluate the time presence and/or the sound level of specific sound source categories using bipolar scales. To the best of our knowledge, only \cite{lavandier2006contribution} relies on objective measurement of those indicators by manually annotating the recordings used as stimuli. This annotation step is hardly manageable for \emph{in-situ} studies such as \cite{ricciardi2015sound} involving a large number of participants. Furthermore, only a pre-imposed restricted set of sound categories are evaluated, preventing experimenters to investigate the influence of more precise or alternative organisations of the sound sources.

@@

\gl{To simulate a verisimilar sound environment, we need to provide the subject with a reasonable and representative number of isolated sounds. This dataset represents the ``\,sonic world\,'', or more precisely a semantic discretization of it in terms of  ``\,sound sources categories\,'', where the term ``\,category\,'' refers to a group of objects (in our case sound sources) considered as equivalent.}

@@

Figure~\ref{datasetexample} shows a subset of the described hierarchical organization. The top classes our taxonomy represent concepts such as ``\,urban transport\,'', encompassing classes of sound sources such as ``\,boat\,'' or  ``\,cars\,'', which are themselves split into sub-classes. At the leaf level, classes are described using sound source names and actions (\textit{car passing, male-voice yelling}). The deeper the class level, the lower the variability between the exemplars lying under the class. In this paper the class levels will be referred to as \emph{semantic levels}, the top classes being at semantic level 0, the first subclasses at semantic level 1 and so on. Leaf classes are collections of recorded isolated sounds.

@@

The sample selection process is designed to rely only on the listening of sounds themselves, in order not to influence subjects with \textit{a priori} associated semantic values. To explore the dataset without any written textual help, we have designed a selection interface where each leaf class is depicted by a point in 2D space, \gl{which was previously tested in an other study} \cite{lafayJAES}. Subjects may only interact with the leaf classes of the datasets but the location of the points in space is constrained by the taxonomy, \textit{i.e.} subclasses belonging to a same class are close to one another. This organization is chosen to facilitate the sound class selection process and allow the subject to quickly apprehend the full range of the dataset. This may also reduce several potential biases of studies relying on verbal data to describe a soundscape: 1) the variable fluency of the subject in the experiment language, 2) the lack of definite terms to fully describe sounds or complex sound environments \cite{guastavino_ideal_2006}, and 3) the high inter-subject variability in the verbal description of a same sound.

@@

Considering a soundscape as a sum of isolated samples is in line with the Auditory Scene Analysis (ASA) theory \cite{bregman1994auditory}, which states that auditory stimuli are decomposed into ``\,auditory objects\,'' called ``\,streams\,'', which may be regarded as sequences of auditory events emitted by putative sound sources \cite{ciocca2007auditory,winkler2009modeling}, and also with recent neuroscience studies suggesting that the primary auditory cortex (A1) produces representations of such ``\,auditory objects\,'' that act as bases for high level cognitive processing \cite{nelken2008neurons}. Thus, the track can be seen as the generative counterpart of the stream, which is rather a cognitive object.\gl{NM comments}

@@

We apply the proposed protocol to the field of urban soundscapes appraisal, by asking a panel of subjects to simulate two antagonistic urban soundscapes:  one ``\,ideal\,'', or pleasant (i) and the other ``\,non-ideal\,'' or unpleasant (ni). The two labels have been chosen to compare the results with those of a previous study addressing ideal urban sound environment representations \cite{guastavino_ideal_2006}. The simulation protocol used in this experiment was previously tested and validated thanks to a pilot study~\cite{soundscape3}.

@@

\begin{enumerate}
\item gain finer knowledge about the pleasantness of each simulated scene,
\item detect possible scene outliers. As the distinction i/ni acts as our reference throughout the analysis, one has to ensure that all the scenes are indeed perceived as ideal or non ideal scenes without ambiguity, \textit{i.e.} the highest average perceptual rating of the ni-scenes has to remain inferior to the lowest average rating of the i-scenes.
\end{enumerate}

The simulation experiment is referred as experiment-1 and the rating experiment as experiment-2. The complete experimental plan is described in Figure~\ref{fig:expLan}.

The simulated scenes as well as their annotations are available on an Archive Data Repository \footnote{\url{https://archive.org/details/soundSimulatedUrbanScene}}. Supplementary information including the experiment webpage and raw data for each experiment can be found on the companion website \footnote{\url{http://www.irccyn.ec-nantes.fr/~lagrange/demonstrations/simScene.html}}. \gl{NM: Pas d'accord commentaires section IV}

@@

483 monophonic urban environmental sounds are collected, including 381 sound events and 102 textures are chosen to build the dataset of isolated sound samples. Among them, 260 events and 72 textures are recorded by the authors using a \textit{Audio Technica AT8035} shotgun microphone (to isolate sounds from their background) connected to a \textit{ZOOM H4n} recorder. The remaining comes from existing sound banks \gl{NM préciser lesquels ?}. All sounds are normalized to the same root mean square (RMS) level.

@@

For ease of reading, we introduce the following terms and concepts that will be used throughout the analysis: 
\begin{description}
\item[$Diversity$:] number of distinct sound classes used in a scene. This number depends on the considered semantic level. For example, considering the two subclasses of the semantic level 2 \textit{car-passing} and \textit{car-starting}, which both belong to the same class \textit{car}, we count 2 distinct sound classes for semantic levels 2 and 3, but only 1 sound class for semantic levels 0 and 1. 
\item[$Density$ ($D$):] logarithm of the number of event samples per chunk of 1 second, averaged over the scene duration. Density may be computed considering all sound classes, or only a subset of them. If nothing is specified, then the density accounts for all classes. \gl{Empirically, we observed that the original density distributions were skewed. The use of the logarithm compression allowed us to reduce this skewness.} 
\item[$Sound level$ ($L$):]  Decibel of the RMS voltage relative to 1 volt, computed every second and averaged over the scene duration. An A-weighting filtering is performed before the RMS computation. Computing other statistical descriptors, namely the minimum, maximum or the 10-90th quantiles, shows that they all present a high correlation with the mean ($r\geq0.85$, $p<0.01$), therefore only the latter is considered.
\end{description}

@@

Subjects are asked to successively create two urban soundscapes, lasting one minute each. The first must be ideal (a soundscape in which the subject would like to live), the second non-ideal (one in which they would try to avoid living). They are asked to mimic a static listener. Subjects are not restricted in their design choices, but are forbidden to create physically impossible situations such as ``\,a dog barking every 10 milliseconds\,''. At the beginning of the experiment, a 20 minutes tutorial is planned in order to familiarize the subjects with the software environment. The experiment is scheduled to last from two to three hours.

@@

Subjects are asked to evaluate the pleasantness of the 80 simulated scenes using a bipolar scale (also known as Likert scale) of 7 points going from -3 (non ideal, very unpleasant) to +3 (ideal, very pleasant). The first ten scenes are used to give the subject an idea of the pleasantness range of the scenes. Those ten scenes are repeated at the end of the experiment and the ratings of their first occurrences are not taken into account in the analysis. Subjects listen to at least the 20 first seconds of the scenes and are free to skip the last 10 seconds at their convenience. Scenes are presented in a randomized order, but the ten calibration scenes are five i-scenes and five ni-scenes. \gl{NM comments: la randomization est déjà précisée.}

@@

@@

Looking at the database criticisms, 28 subjects stated that they could not find one or several desired sounds, with a maximum of 4 sounds for one subject. From all the missing sounds we identified 26 sound classes at different semantic levels. Among those classes, 16 were effectively present in the database, 1 referred to musical sounds which we had chosen to exclude from the database, and only 9 were effectively absent. Except for one sound class (``\,stroller pushchair\,''), each of the 16 classes which were reported as missing but were present in the database were used by at least one other subject. Among the 9 truly missing ones, we find very specific sounds as ``\,sport car\,'' or ``\,teenager voice\,''. We believe that those results show that the database diversity was sufficient for the purpose of the study.

@@

\begin{figure}[t]
\begin{center}
\includegraphics[width=.4\paperwidth]{../gfxMatlab/xp2_1.eps}
\caption{\label{xp2_1} Boxplot of the averaged ratings given by the subject considering (a) scenes as observations, (b) scenes as observations without scene outliers and (c) subjects as observations without scene outliers.} 
\end{center}
\end{figure}

@@

\begin{figure}[t!]
\centering
\includegraphics[width=.4\paperwidth]{../gfxMatlab/xp1_deSoLv_1.eps} 
\caption{\label{fig:xp1_deSoLv_1}  Boxplot of (a) the scene densities ($D$) and (b) the scene sound levels ($L$). Scatter plot of (c) $D$ \emph{vs} ratings and (d) $L$ \emph{vs} rating}
\end{figure}

@@

\begin{figure}[t!]
\centering
\includegraphics[width=.4\paperwidth]{../gfxMatlab/xp1_div_1.eps} 
\caption{\label{fig:analyseDensitySnrGlobal3} Diversity across semantic levels for event and texture classes.}
\end{figure}

@@

To study the ability of those descriptors to characterize the appraisal, we compute the correlations between the ratings, $D$ and $L$, considering separately the i- and ni-scenes, as wel as all the scenes altogether. Results are shown in Table~\ref{tab:numGlobal} and in Figures~\ref{fig:xp1_deSoLv_1}.c and~\ref{fig:xp1_deSoLv_1}.d; \gl{NM comments: je ne pense pas qu'une MANOVA s'utilise comme cela}

@@

Averaged diversities are shown on Figure~\ref{fig:analyseDensitySnrGlobal3}. Except for semantic level 0, the diversity is significantly higher for the event classes of the ni-scenes (levels 1, 2 and 3: $p<0.01$;), revealing that ni-scenes are composed of a larger variety of sound events than i-scenes. Diversity remains constant across semantic levels for texture classes of i- and ni-scenes, with no statistical differences. The correlations computed between the event diversities and the averaged pleasantness ratings remain weak or moderate ($r<0.45$ \gl{NM comments: non}, $p<0.01$), irrespective of the considered semantic level. \gl{NM comments: non} If we look at the two type of scene separately, no correlation is observed. \gl{Although the diversity varies between the i- and ni-scenes, this last result shows that it is not a good indicator to precisely characterize the pleasantness.}

>>>>

\newcommand{\gl}[1]{\textcolor{blue}{GL : #1}}
\newcommand{\nm}[1]{\textcolor{red}{nm : #1}}
\newcommand{\ml}[1]{\textcolor{magenta}{ML : #1}}
\newcommand{\mr}[1]{\textcolor{green}{MR : #1}}

@@

As a case study to demonstrate the interest of that new approach, we focus in this paper on the issue of urban soundscape pleasantness as tackled in \cite{guastavino_ideal_2006} \nm{dans [14], il est plutot question de ideal / non ideal urban soundscapes ..... }.

@@

We first give an overview of the tools used in soundscape perception studies, and justify the benefits of conducting this kind of simulation experiment as a data generation process. We then describe the simulation process, and present the results of the case study. The latter consists of two experiments: 1) a simulation experiment where subjects generate both ideal and non-ideal urban sonic environments, and 2) a pleasantness evaluation of the simulated scenes using a 7 points bipolar semantic scale. \nm{en revoyant cette description, je me dis qu'il y a pt-etre un petit probleme methodologique dans la mesure ou l'on contruit les scenes selon un attribut semantique (l'idealite) et on les evalue sur une autre echelle (l'agrement) .....
Je pense qu'il faudra bien argumenter c point plus bas dans le texte}

Results show that \nm{Cette section sur les resultats est etonnemment placee, et a priori, de maniere pas tres orthodoxe. Je verrai plutot cette synthese des resultats dans l'abstract ou dans la conclusion ....} 1) outcomes identified in the literature \cite{guastavino_categorization_2007} can be achieved by this approach, providing evidence that it has solid grounding, 2) sound markers that greatly affect the pleasantness of the soundscape can be identified with a good level of categorical detail, and 3) that physical descriptors computed specifically on the sound markers better characterize the appraisal than global descriptors.

@@

Two approaches, named in this paper \textit{categorical} and \textit{dimensional}, are commonly used to gain knowledge on soundscape perception \nm{this is not really specific to soundscape ... but a global approach of perception mechanisms. We can find this distinction also, for instance, in the works about musical timbre}.

The \textit{categorical} approach investigates both the existence of 1) soundscape categories and 2) soundscape components (categories of sound sources) using both sorting \cite{maffiolo_caracterisation_1999, guastavino_categorization_2007} and descriptive \cite{axelsson2005soundscape, raimbault_urban_2005, raimbault2006qualitative, guastavino_ideal_2006} tasks (questionnaires or interviews). To identify categories, those methods usually require psycho-linguistic and lexical analyses. Their results tend to show that soundscape perception relies upon the identification and assessment of the sound sources which compose the soundscape \cite{dubois2006cognitive}. Maffiolo \cite{maffiolo_caracterisation_1999} shows the existence of two generic soundscape categories, namely the ``\,event sequences\,'' and the ``\,amorphous sequences\,''. \gl{This distinction is also observed in \cite{guastavino_ideal_2006}}. Guastavino \cite{guastavino_categorization_2007} as well as Niessen \emph{et al.} \cite{niessen_categories_2010}  show the prominence of specific sound source categories in soundscape perception: ``\,natural\,'', ``\,human\,'', ``\,traffic\,'' and ``\,mechanical\,''. Other studies not using the \textit{categorical} approach reveal that sound source identification underlines the perception of others factors, for example Kuwano \emph{et al.} \cite{kuwano_memory_2003-1} show that overall judgment of loudness of a sound environment sequence is not statistically different from average instantaneous judgment of the recalled sound events of the sequence \nm{Cet exemple n'est pas completement clair pour moi :
certes, la loudness est bien une dimension continue (et donc en cela rentre bien dans ta terminologie de depart) mais je ne vois pas trop en quoi elle a un rapport avec la question de l'identification des sources ...
a preciser, je pense.}.  As reported by \cite{davies_perception_2013}, while the categorical approach does help to understand how a soundscape is composed, it does not provide information on how soundscape components are perceived and influence the qualitative evaluation.

Instead of asking subjects to break soundscapes down into categories, the \textit{dimensional} approach makes use of bipolar semantic scales to describe the soundscape either in a holistic way \cite{jeon2013soundwalk, kang_semantic_2010, cain_development_2013, torija2013application} or by considering the influence of specific sound sources \cite{guyot2005urban, lavandier2006contribution, ricciardi2015sound}. From those descriptors, it is possible to use dimensional analysis techniques such as Principal Component Analysis to find relevant perceptual axes \cite{cain_development_2013}, or to connect those perceptive outcomes with objective acoustics measurements using various statistical analysis or modelling techniques such as multiple linear regression \cite{lavandier2006contribution,ricciardi2015sound} or artificial neural network \cite{yu2009modeling}. Several semantic scales show promising results when it comes to modeling perceived affective quality, namely ``\,Calmness\,'' and ``\,Vibrancy\,'' \cite{cain_development_2013, axelsson2010principal}, or ``\,Pleasantness\,'' and ``\,Eventfulness\,'' \cite{axelsson2010principal}; but it has been shown that others context dependent (``\,Appropriateness\,'',``\,Familiarity\,'') or music related (``\,Music-likeness\,'' ) scales provide useful complementary information on soundscape perception \cite{aletta2016soundscape}. In addition, several studies have shown that multiple factors such as physical, behavioral, social, demographical, and psychological are involved when evaluating acoustic comfort \cite{marquis2005noise, yu2009modeling}. \gl{In particular} \cite{yu2010factors} shows that age and education level significantly influence the sound preference \nm{Je pense qu'il faut etre un peu plus critique sur ces questions la (et celles developpees dans la phrase precedente) et sur ce sujet qu'il faut absolument citer le papier de synthese de C. Marquis-Favre :} \gl{Attention \cite{marquis2005noise} est une review ...}.

\gl{Recent} studies investigating the influence of sound sources in soundscape perception show that global sound level (perceived or measured) cannot entirely explain the quality of a sonic environment. \gl{voir remarque precedente; sur cette question, je privilegierais plutot un artcile de revue (AAA), voire un bouquin (Kang,2006) qu'un proc. de conf. ...} \cite{guyot2005urban} shows that a rise in soundscape quality between two street soundscapes having a similar time-ratio of traffic sounds and a similar global sound level may result from the addition of a positively connotated sound source, e.g. belonging to the ``\,human\,'' category \nm{mais pas que de la categorie human; cf. le resultat que tu evoques  plus bas de [13] avec les sources type "public transport" egalement connotes plus positif que des "vehicules legers" du fait de leur caractere ecologique et collectif plus important que la voiture individuelle} \gl{eg means exempli gratia (par exemple), de plus nous montrons que bien  que les bus soient acceptés (presence dans les i-scenes), ils ne sont pas caracteristiques d'un environnement ideal (presence dans les ni-scenes)}. Using multiple linear regression to predict unpleasantness from descriptors characterizing global sound level as well as time presence and sound level of each source, \cite{lavandier2006contribution} shows that the percentage of explained variance increases when the source descriptors are added to the global sound level. Similar findings are made in \cite{ricciardi2015sound}, showing that a combined use of global and source specific descriptors allows to build more relevant model of sound quality. Lastly, \cite{lavandier2006contribution} shows that global sound level is not always appropriate to characterize the appraisal, especially in the case of positively rated soundscapes such as \textit{park} or \textit{market}.

In order to be able to study the relative perceptual importance of each sound source, it is beneficial to first gain knowledge about the presence and level of those sound sources within \nm{inside ?} the scene. Acquiring such data may be a serious challenge. From most of the studies cited above \cite{guyot2005urban, lavandier2006contribution, ricciardi2015sound}, subjects are asked to evaluate the time presence and/or the sound level of specific sound source categories using bipolar scales. To the best of our knowledge, only \cite{lavandier2006contribution} relies on objective measurement of those indicators by manually annotating the recordings used as stimuli. This annotation step is hardly manageable for \emph{in-situ} studies such as \cite{ricciardi2015sound} involving a large number of participants. Furthermore, only a pre-imposed restricted set of sound categories are evaluated, preventing experimenters to investigate the influence of more precise or alternative organisations of the sound sources.

@@

\gl{To simulate a verisimilar sound environment, we need to provide the subject with a reasonable and representative number of isolated sounds. This dataset represents the ``\,sonic world\,'', or more precisely a semantic discretization of it in terms of  ``\,sound sources categories\,''. The term ``\,category\,'' refers to a group of objects (in our case sound sources) considered as equivalent.}

@@

Figure~\ref{datasetexample} shows a subset of the described hierarchical organization. The top classes of our taxonomy represent concepts such as ``\,urban transport\,'', encompassing classes of sound sources such as ``\,boat\,'' or  ``\,cars\,'', which are themselves split into sub-classes. At the leaf level, classes are described using sound source names and actions (\textit{car passing, male-voice yelling}). The deeper the class level, the lower the variability between the exemplars lying under the class. In this paper the class levels will be referred to as \emph{semantic levels}, the top classes being at semantic level 0, the first subclasses at semantic level 1 and so on. Leaf classes are collections of recorded isolated sounds.

@@

The sample selection process is designed to rely only on the listening of sounds themselves, in order not to influence subjects with \textit{a priori} associated semantic values. To explore the dataset without any written textual help, we have designed a selection interface where each leaf class is depicted by a point in 2D space. \gl{The interface performances are extensively tested in the following study} \cite{lafayJAES}. Subjects may only interact with the leaf classes of the datasets but the location of the points in space is constrained by the taxonomy, \textit{i.e.} subclasses belonging to a same class are close to one another. This organization is chosen to facilitate the sound class selection process and allow the subject to quickly apprehend the full range of the dataset. This may also reduce several potential biases of studies relying on verbal data to describe a soundscape: 1) the variable fluency of the subject in the experiment language, 2) the lack of definite terms to fully describe sounds or complex sound environments \cite{guastavino_ideal_2006}, and 3) the high inter-subject variability in the verbal description of a same sound.

@@

Considering a soundscape as a sum of isolated samples is in line with the Auditory Scene Analysis (ASA) theory \cite{bregman1994auditory}, which states that auditory stimuli are decomposed into ``\,auditory objects\,'' called ``\,streams\,'', which may be regarded as sequences of auditory events emitted by putative sound sources \cite{ciocca2007auditory,winkler2009modeling}, and also with recent neuroscience studies suggesting that the primary auditory cortex (A1) produces representations of such ``\,auditory objects\,'' that act as bases for high level cognitive processing \cite{nelken2008neurons}. Thus, the track can be seen as the generative counterpart of the stream, which is rather a cognitive object.\nm{ce paragraphe est interessant mais est, a mon avis, mal place; il legitime la structure du simulateur, je le mettrais donc plutot dans le §3.1 en le developpant pt-etre un peu plus notamment les aspects neurosciences ...}

@@

We apply the proposed protocol to the field of urban soundscapes appraisal, by asking a panel of subjects to simulate two antagonistic urban soundscapes:  one ``\,ideal\,'', or pleasant (i) and the other ``\,non-ideal\,'' or unpleasant (ni) \nm{voir ma remarque en intro. sur la concordance ideal/pleasant ...}. The two labels have been chosen to compare the results with those of a previous study addressing ideal urban sound environment representations \cite{guastavino_ideal_2006}. The simulation protocol used in this experiment was previously tested and validated thanks to a pilot study~\cite{soundscape3}.

@@

\begin{enumerate}
\item gain finer knowledge about the pleasantness of each simulated scene, \nm{trop vague a mon sens : a developper}
\item detect possible scene outliers. As the distinction i/ni \nm{cet encodage n'a  jamais ete defini avant - meme si on devine bien ce qu'il veut dire .... ;)} \gl{cf. plus haut} acts as our reference throughout the analysis, one has to ensure that all the scenes are indeed perceived as ideal or non ideal \nm{il ya confusion entre l'evalution ideal/non ideal et sur l'echelle de pleasantness .... est-ce la meme expe. ? a clarifier} scenes without ambiguity, \textit{i.e.} the highest average perceptual rating of the ni-scenes has to remain inferior to the lowest average rating of the i-scenes.
\end{enumerate}

The simulation experiment is referred as experiment-1 and the rating experiment as experiment-2. The complete experimental plan is described in Figure~\ref{fig:expLan}. \nm{Du coup, cette info; doit etre mise avant dans le paragraphe histoire d'evacuer toutes les confusions (cf. mes remarques plus haut)} \gl{non, je pense qu'à moins de deux paragraphes d'intervales, les lecteurs vont s'y retrouver.}

The simulated scenes as well as their annotations are available on an Archive Data Repository \footnote{\url{https://archive.org/details/soundSimulatedUrbanScene}}. Supplementary information including the experiment webpage and raw data for each experiment can be found on the companion website \footnote{\url{http://www.irccyn.ec-nantes.fr/~lagrange/demonstrations/simScene.html}}.

@@

483 monophonic urban environmental sounds are collected, including 381 sound events and 102 textures are chosen to build the dataset of isolated sound samples. Among them, 260 events and 72 textures are recorded by the authors using a \textit{Audio Technica AT8035} shotgun microphone (to isolate sounds from their background) connected to a \textit{ZOOM H4n} recorder. The remaining comes from existing sound banks \nm{preciser lesquelles ?} \gl{non}. All sounds are normalized to the same root mean square (RMS) level.

@@

@@

Subjects are asked to successively create two urban soundscapes, lasting one minute each. The first must be ideal (a soundscape in which the subject would like to live), the second non-ideal (one in which they would try to avoid living) \nm{reprendre la formulation exacte de la consigne} \gl{C'est la formulation, bien que cette dernière était transmise oralement}. They are asked to mimic a static listener. Subjects are not restricted in their design choices, but are forbidden to create physically impossible situations such as ``\,a dog barking every 10 milliseconds\,''. At the beginning of the experiment, a 20 minutes tutorial is planned in order to familiarize the subjects with the software environment. The experiment is scheduled to last from two to three hours.

@@

Subjects are asked to evaluate the pleasantness of the 80 simulated scenes using a bipolar scale (also known as Likert scale) of 7 points going from -3 (non ideal, very unpleasant) \nm{a nouveau cet amalgame des 2 termes risque d'etre critiquee et doit etre argumenter et expliciter, je pense.} to +3 (ideal, very pleasant). The first ten scenes are used to give the subject an idea of the pleasantness range of the scenes. Those ten scenes are repeated at the end of the experiment and the ratings of their first occurrences are not taken into account in the analysis. Subjects listen to at least the 20 first seconds of the scenes and are free to skip the last 10 seconds at their convenience. Scenes are presented in a randomized order, but the ten calibration scenes are five i-scenes and five ni-scenes. \nm{also randomized in the presentation along subjects ?} \gl{a vérifier}

@@

For ease of reading, we introduce the following terms and concepts that will be used throughout the analysis: 
\begin{description}
\item[$Diversity$:] number of distinct sound classes used in a scene. This number depends on the considered semantic level. For example, considering the two subclasses of the semantic level 2 \textit{car-passing} and \textit{car-starting}, which both belong to the same class \textit{car}, we count 2 distinct sound classes for semantic levels 2 and 3, but only 1 sound class for semantic levels 0 and 1. 
\item[$Density$ ($D$):] logarithm of the number of event samples per chunk of 1 second, averaged over the scene duration. Density may be computed considering all sound classes, or only a subset of them. If nothing is specified, then the density accounts for all classes. \gl{Empirically, we observed that the original distributions were skewed. The use of the logarithm compression allowed us to reduce this skewness.} 
\item[$Sound level$ ($L$):]  Decibel of the RMS voltage relative to 1 volt, computed every second and averaged over the scene duration. \nm{est-ce que cette reference correspond a une quelconque calibration du microphone
de prise de son (p. ex., X dB a 1 m.) ? Si oui, donner plus de details et expliquer comment tu as fait pour les samples issus de base de donnees - dont on ne connait pas a priori ces elements de calibration .....} \gl{non} An A-weighting filtering is performed before the RMS computation. Computing other statistical descriptors, namely the minimum, maximum or the 10-90th quantiles, shows that they all present a high correlation with the mean ($r\geq0.85$, $p<0.01$), therefore only the latter is considered.
\end{description}

@@

Looking at the database criticisms, 28 subjects stated that they could not find one or several desired sounds, with a maximum of 4 sounds for one subject. From all the missing sounds we identified 26 sound classes at different semantic levels. Among those classes, 16 were effectively present in the database, 1 referred to musical sounds which we had chosen to exclude from the database, and only 9 were effectively absent. Except for one sound class (``\,stroller pushchair\,''), each of the 16 classes which were reported as missing but were present in the database were used by at least one other subject. Among the 9 truly missing ones, we find very specific sounds as ``\,sport car\,'' or ``\,teenager voice\,''. We believe \nm{n'est-ce pas unpeu faible comme argumentaire ....? Pt-etre mettre les resultats en \% pour les rendre plus facilement intrerpretables ?} that those results show that the database diversity was sufficient for the purpose of the study.

@@

\begin{figure}[t]
\begin{center}
\includegraphics[width=\columnwidth]{../gfxMatlab/xp2_1.eps}
\caption{\label{xp2_1} Boxplot of the averaged ratings given by the subject considering (a) scenes as observations, (b) scenes as observations without scene outliers and (c) subjects as observations without scene outliers. \nm{les figures sont petites et le chiox des couleurs contribue a leur faible lisibilite} \gl{?}} 
\end{center}
\end{figure}

@@

\begin{figure}[t!]
\centering
\includegraphics[width=\columnwidth]{../gfxMatlab/xp1_deSoLv_1.eps} 
\caption{\label{fig:xp1_deSoLv_1}  Boxplot of (a) the scene densities ($D$) and (b) the scene sound levels ($L$). Scatter plot of (c) $D$ \emph{vs} ratings and (d) $L$ \emph{vs} rating}
\end{figure}

@@

\begin{figure}[t!]
\centering
\includegraphics[width=.4\paperwidth]{../gfxMatlab/xp1_div_1.eps} 
\caption{\label{fig:analyseDensitySnrGlobal3} Diversity across semantic levels for event and texture classes. \nm{il faut mieux differencier les donnees provenant du corpus event et du corpus texture}}
\end{figure}

@@

To study the ability of those descriptors to characterize the appraisal, we compute the correlations between the ratings, $D$ and $L$, considering separately the i- and ni-scenes, as wel as all the scenes altogether. Results are shown in Table~\ref{tab:numGlobal} and in Figures~\ref{fig:xp1_deSoLv_1}.c and~\ref{fig:xp1_deSoLv_1}.d; \nm{pour rendre compte de cet effet, ne peut-on pas egalement calculer effectuer une MANOVA qui renseignera sur l'interaction de 1er ordre entre la variable dependante (ratings) et les variables independantes / facteurs expe. (density et Leq) ?} \gl{NM comments: je ne pense pas qu'une MANOVA s'utilise comme cela, comme toute analyse de variance, la variable independante est categorielle. De plus dans une MANOVA, le M de multivarié s'applique aux variales dépendantes, la proposition relève de la two way anova (mais le rating étant continu cela ne marche pas). Il faudrait faire une regression multiple avec terme d'interaction P=A+B+A*B, et interpréter A*B...}

@@

Averaged diversities are shown on Figure~\ref{fig:analyseDensitySnrGlobal3}. Except for semantic level 0, the diversity is significantly higher for the event classes of the ni-scenes (levels 1, 2 and 3: $p<0.01$;), revealing that ni-scenes are composed of a larger variety of sound events than i-scenes. Diversity remains constant across semantic levels for texture classes of i- and ni-scenes, with no statistical differences. \nm{qu'en est-il pour les classes event ? Est-ce que les differences observees sur fig. 7 (a partir de semantic level 1) sont statist. siginificatives ?} \gl{cf. phrase précédente} The correlations computed between the event diversities and the averaged pleasantness ratings remain weak or moderate ($r<0.45$ \nm{r = 0.45 ?} \gl{NM comments: non, toutes les correlations sont inf à .45}, $p<0.01$), irrespective of the considered semantic level.  If we look at the two types of scene separately, no correlation is observed. \gl{Although the diversity varies between the i- and ni-scenes, this last result shows that it is not a good indicator to precisely characterize the pleasantness.}
